---
title: "EDS232_lab1"
author: "Joe DeCesaro"
date: "1/3/2022"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = T)
```

## Load Packages
```{r}
# load packages, installing if missing
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  dismo, dplyr, DT, ggplot2, here, htmltools, leaflet, mapview, purrr, raster, readr, rgbif, rgdal, rJava, sdmpredictors, sf, spocc, tidyr)
select <- dplyr::select # overwrite raster::select
options(readr.show_col_types = FALSE)

# set random seed for reproducibility
set.seed(42)

# directory to store data
dir_data <- here("data/sdm")
dir.create(dir_data, showWarnings = F, recursive = T)
```

## Get Species Observations
```{r}
obs_csv <- file.path(dir_data, "obs.csv")
obs_geo <- file.path(dir_data, "obs.geojson")
redo    <- FALSE

if (!file.exists(obs_geo) | redo){
  # get species occurrence data from GBIF with coordinates
  (res <- spocc::occ(
    query = 'Enhydra lutris', 
    from = 'gbif', has_coords = T,
    limit = 10000))
  
  # extract data frame from result
  df <- res$gbif$data[[1]] 
  readr::write_csv(df, obs_csv)
  
  # convert to points of observation from lon/lat columns in data frame
  obs <- df %>% 
    sf::st_as_sf(
      coords = c("longitude", "latitude"),
      crs = st_crs(4326)) %>% 
    select(prov, key) # save space (joinable from obs_csv)
  sf::write_sf(obs, obs_geo, delete_dsn=T)
}
```

### Extract data from the result
```{r}
obs <- sf::read_sf(obs_geo)
nrow(obs) # number of rows
```

### Show Points on Map
```{r}
# show points on map
mapview::mapview(obs, map.types = "Esri.OceanBasemap")
```

## Question 1: How many observations total are in GBIF for your species?
13,716

## Question 2: Do you see any odd observations, like marine species on land or vice versa?
It looks like there are a few otters in the North Atlantic Ocean which does not make sense for this species. Also, there is one sighting in the Arctic, north of Alaska and one in the middle of Alaska. All of these will be accounted for below. The points in the Atlantic appear to be missing a 1 at the beginning on their longitude coordinate. All points were removed using there index number in the `obs` dataframe.

```{r}
obs <- obs[-c(1524, 
              2102, 
              2526, 
              4035,
              4044,
              6806,
              6850, 
              7337), ] 
```

## Get Environmental Data
### Presensce
```{r}
dir_env <- file.path(dir_data, "env")

# set a default data directory
options(sdmpredictors_datadir = dir_env)

# choosing marine
env_datasets <- sdmpredictors::list_datasets(terrestrial = FALSE, marine = TRUE)

# show table of datasets
env_datasets %>% 
  select(dataset_code, description, citation) %>% 
  DT::datatable()

# choose datasets for a vector
env_datasets_vec <- c("Bio-ORACLE", "MARSPEC")

# get layers
env_layers <- sdmpredictors::list_layers(env_datasets_vec)
DT::datatable(env_layers)

# choose layers after some inspection and perhaps consulting literature 
env_layers_vec <- c("BO_sstmean", "BO_bathymean", "MS_biogeo05_dist_shore_5m", "BO2_carbonphytomean_bdmean", "BO_parmean", "BO_ph")

# get layers
env_stack <- load_layers(env_layers_vec, rasterstack = FALSE)

# interactive plot layers, hiding all but first (select others)
# mapview(env_stack, hide = T) # makes the html too big for Github
plot(env_stack, nc=2)
```
**The above line `plot(env_stack, nc=2)` does not work, getting an error of "Error in xy.coords(x, y, xlabel, ylabel, log) : 'x' is a list, but does not have components 'x' and 'y'"** 

### 2.4.2 Pseudo-Absence
```{r make absence pts}
absence_geo <- file.path(dir_data, "absence.geojson")
pts_geo     <- file.path(dir_data, "pts.geojson")
pts_env_csv <- file.path(dir_data, "pts_env.csv")
redo <- FALSE

if (!file.exists(absence_geo) | redo){
  # get raster count of observations
  r_obs <- rasterize(
    sf::as_Spatial(obs), env_stack[[1]], field=1, fun='count')
  
  # show map
  # mapview(obs) + 
  #   mapview(r_obs)
  
  # create mask for 
  r_mask <- mask(env_stack[[1]] > -Inf, r_obs, inverse=T)
  
  # generate random points inside mask
  absence <- dismo::randomPoints(r_mask, nrow(obs)) %>% 
    as_tibble() %>% 
    st_as_sf(coords = c("x", "y"), crs = 4326)
  
  write_sf(absence, absence_geo, delete_dsn=T)
}
absence <- read_sf(absence_geo)

# show map of presence, ie obs, and absence
mapview(obs, col.regions = "green") + 
  mapview(absence, col.regions = "gray")
```


```{r make absence pts}
### Code breaks here
redo <- TRUE
if (!file.exists(pts_env_csv) | redo){

  # combine presence and absence into single set of labeled points 
  pts <- rbind(
    obs %>% 
      mutate(
        present = 1) %>% 
      select(present, key),
    absence %>% 
      mutate(
        present = 0,
        key     = NA)) %>% 
    mutate(
      ID = 1:n()) %>% 
    relocate(ID)
  write_sf(pts, pts_geo, delete_dsn=T)

  # extract raster values for points
  pts_env <- raster::extract(env_stack, as_Spatial(pts), df=TRUE) %>% 
    tibble() %>% 
    # join present and geometry columns to raster value results for points
    left_join(
      pts %>% 
        select(ID, present),
      by = "ID") %>% 
    relocate(present, .after = ID) %>% 
    # extract lon, lat as single columns
    mutate(
      #present = factor(present),
      lon = st_coordinates(geometry)[,1],
      lat = st_coordinates(geometry)[,2]) %>% 
    select(-geometry)
  write_csv(pts_env, pts_env_csv)
}
pts_env <- read_csv(pts_env_csv)
```


```{r make absence pts}
pts_env %>% 
  # show first 10 presence, last 10 absence
  slice(c(1:10, (nrow(pts_env)-9):nrow(pts_env))) %>% 
  DT::datatable(
    rownames = F,
    options = list(
      dom = "t",
      pageLength = 20))
```


## Term Plots
```{r plot terms}
pts_env %>% 
  select(-ID) %>% 
  mutate(
    present = factor(present)) %>% 
  pivot_longer(-present) %>% 
  ggplot() +
  geom_density(aes(x = value, fill = present)) + 
  scale_fill_manual(values = alpha(c("gray", "green"), 0.5)) +
  scale_x_continuous(expand=c(0,0)) +
  scale_y_continuous(expand=c(0,0)) +
  theme_bw() + 
  facet_wrap(~name, scales = "free") +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0))
```
