---
title: 'Lab 1b. Species Distribution Modeling'
author: "Joe DeCesaro"
date: "1/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Explore (cont'd)

```{r load pkgs data}
librarian::shelf(
  DT, dplyr, dismo, GGally, here, readr, tidyr)
select <- dplyr::select # overwrite raster::select
options(readr.show_col_types = F)

dir_data    <- here("data/sdm")
pts_env_csv <- file.path(dir_data, "pts_env.csv")

pts_env <- read_csv(pts_env_csv)
nrow(pts_env)

datatable(pts_env, rownames = F)
```

Let's look at a pairs plot (using [GGally::ggpairs()](http://ggobi.github.io/ggally/articles/ggpairs.html)) to show correlations between variables. 

```{r ggpairs, fig.cap="Pairs plot with `present` color coded.", fig.width=11, fig.height=11, eval=T}
GGally::ggpairs(
  select(pts_env, -ID),
  aes(color = factor(present)))
```

# Logistic Regression

## Setup Data
Let's setup a data frame with only the data we want to model by:

```{r setup d}
# setup model data
d <- pts_env %>% 
  select(-ID) %>%  # remove terms we don't want to model
  tidyr::drop_na() # drop the rows with NA values
nrow(d)
```

## Linear Model
Let's start as simply as possible with a linear model `lm()` on multiple predictors `X` to predict presence `y` using a simpler workflow.

```{r fit lm}
# fit a linear model
mdl <- lm(present ~ ., data = d)
summary(mdl)
```


```{r fit lm}
y_predict <- predict(mdl, d, type="response")
y_true    <- d$present

range(y_predict)
range(y_true)
```

The problem with these predictions is that it ranges outside the possible values of present `1` and absent `0`. (Later we'll deal with converting values within this range to either `1` or `0` by applying a cutoff value; i.e. any values > 0.5 become 1 and below become 0.)

## Generalized Linear Model

To solve this problem of constraining the response term to being between the two possible values, i.e. the **probability** $p$ of being one or the other possible $y$ values, we'll apply the logistic transformation on the response term.

$$
logit(p_i) = \log_{e}\left( \frac{p_i}{1-p_i} \right)
$$
We can expand the expansion of the predicted term, i.e. the probability $p$ of being either $y$, with all possible predictors $X$ whereby each coefficient $b$ gets multiplied by the value of $x$:

$$
\log_{e}\left( \frac{p_i}{1-p_i} \right) = b_0 + b_1 x_{1,i} + b_2 x_{2,i} + \cdots + b_k x_{k,i}
$$

```{r fit glm}
# fit a generalized linear model with a binomial logit link function
mdl <- glm(present ~ ., family = binomial(link="logit"), data = d)
summary(mdl)
```


```{r fit glm}
y_predict <- predict(mdl, d, type="response")

range(y_predict)
```

Excellent, our response is now constrained between 0 and 1. Next, let's look at the term plots to see the relationship between predictor and response.

```{r termplots glm}
# show term plots
termplot(mdl, partial.resid = TRUE, se = TRUE, main = F, ylim="free")
```

## Generalized Additive Model

With a generalized additive model we can add "wiggle" to the relationship between predictor and response by introducing smooth `s()` terms.

```{r fit gam}
librarian::shelf(mgcv)

# fit a generalized additive model with smooth predictors
mdl <- mgcv::gam(
  formula = present ~ s(WC_alt) + s(WC_bio1) + 
    s(WC_bio2) + s(WC_bio4) + s(WC_tmean1) + s(ER_tri) + s(ER_topoWet) + s(lon) + s(lat), 
  family = binomial, data = d)
summary(mdl)
```


```{r fit gam}
# show term plots
plot(mdl, scale=0)
```
- **Question**: Which GAM environmental variables, and even range of values, seem to contribute most towards presence (above 0 response) versus absence (below 0 response)?
  - Presence:
    - "WC_tmean1" (Mean temperature (January)): Possibly for >0 and <-30 degrees but it seems to have a wide variance (dotted lines are not close)
    - Longitude: Possibly from -150 > -120 but it does not seem strong
    - "WC_bio4" (Temperature seasonality): Possibly between 70-110
  - Absence:
    - "WC_bio1" (Annual mean temperature): >10 degrees
    - "WC_bio4" (Temperature seasonality): <70 and >110 
    - Longitude: Possibly from < -150 but it does not seem strong

  Overall it seems like there are more variables that are better predictors for absence rather than presence but that could just be from my inexperience to read these charts.

## Maxent (Maximum Entropy)

Maxent is probably the most commonly used species distribution model ([Elith 2011](http://dx.doi.org/10.1111/j.1472-4642.2010.00725.x)) since it performs well with few input data points, only requires presence points (and samples background for comparison) and is easy to use with a Java graphical user interface (GUI).

```{r fit maxent, echo=TRUE}
# load extra packages
librarian::shelf(
  maptools, sf)

mdl_maxent_rds <- file.path(dir_data, "mdl_maxent.rds")

# show version of maxent
if (!interactive())
  maxent()
```


```{r fit maxent, echo=TRUE}
# get environmental rasters
# NOTE: the first part of Lab 1. SDM - Explore got updated to write this clipped environmental raster stack
env_stack_grd <- file.path(dir_data, "env_stack.grd")
env_stack <- stack(env_stack_grd)
plot(env_stack, nc=2)
```

```{r fit maxent, echo=TRUE}
# get presence-only observation points (maxent extracts raster values for you)
obs_geo <- file.path(dir_data, "obs.geojson")
obs_sp <- read_sf(obs_geo) %>% 
  sf::as_Spatial() # maxent prefers sp::SpatialPoints over newer sf::sf class
```


```{r fit maxent, echo=TRUE}
# fit a maximum entropy model
if (!file.exists(mdl_maxent_rds)){
  mdl <- maxent(env_stack, obs_sp)
  readr::write_rds(mdl, mdl_maxent_rds)
}
mdl <- read_rds(mdl_maxent_rds)
```


```{r fit maxent, echo=TRUE}
# plot variable contributions per predictor
plot(mdl)
```


```{r fit maxent, echo=TRUE}
# plot term plots
response(mdl)
```
- **Question**: Which Maxent environmental variables, and even range of values, seem to contribute most towards presence (closer to 1 response) and how might this differ from the GAM results?
  - "WC_bio1": Temps < -5 degrees
  - "WC_tmean1": Temps > 10 degrees
  - "ER_tri": Values > 300
  - "ER_topoWet": Values < 0
  The Maxent environmental variables add WC_bio1, ER_tri, and ER_topoWet to presence compared to the GAM results and takes away the WC_bio4.

```{r fit maxent, echo=TRUE}
# predict
y_predict <- predict(env_stack, mdl) #, ext=ext, progress='')

plot(y_predict, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```

