---
title: 'Lab 1c. Species Distribution Modeling'
author: "Joe DeCesaro"
date: "1/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r setup}
# global knitr chunk options
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE)

# load packages
librarian::shelf(
  caret,       # m: modeling framework
  dplyr, ggplot2 ,here, readr, 
  pdp,         # X: partial dependence plots
  ranger,      # m: random forest modeling
  rpart,       # m: recursive partition modeling
  rpart.plot,  # m: recursive partition plotting
  rsample,     # d: split train/test data
  skimr,       # d: skim summarize data table
  vip)         # X: variable importance
```


```{r setup}
# options
options(
  scipen = 999,
  readr.show_col_types = F)
set.seed(42)

# graphical theme
ggplot2::theme_set(ggplot2::theme_light())

# paths
dir_data    <- here("data/sdm")
pts_env_csv <- file.path(dir_data, "pts_env.csv")

# read data
pts_env <- read_csv(pts_env_csv)
d <- pts_env %>% 
  select(-ID) %>%                   # not used as a predictor x
  mutate(
    present = factor(present)) %>%  # categorical response
  na.omit()                         # drop rows with NA
skim(d)
```

## Split data into training and testing

```{r dt-data-prereq, echo=TRUE}
# create training set with 80% of full data
d_split  <- rsample::initial_split(d, prop = 0.8, strata = "present")
d_train  <- rsample::training(d_split)

# show number of rows present is 0 vs 1
table(d$present)
table(d_train$present)
```

# Decision Trees

## Partition, depth=1
```{r rpart-stump, echo=TRUE, fig.width=4, fig.height=3, fig.show='hold', fig.cap="Decision tree illustrating the single split on feature x (left).", out.width="48%"}
# run decision stump model
mdl <- rpart(
  present ~ ., data = d_train, 
  control = list(
    cp = 0, minbucket = 5, maxdepth = 1))
mdl

# plot tree 
par(mar = c(1, 1, 1, 1))
rpart.plot(mdl)
```

## Partition, depth=default

```{r rpart-default, echo=TRUE, fig.width=4, fig.height=3, fig.show='hold', fig.cap="Decision tree $present$ classification.", out.width="48%"}
# decision tree with defaults
mdl <- rpart(present ~ ., data = d_train)
mdl
rpart.plot(mdl)

# plot complexity parameter
plotcp(mdl)

# rpart cross validation results
mdl$cptable
```
**Question**: Based on the complexity plot threshold, what size of tree is recommended?
I believe it is recommending a tree of size 5.

## Feature interpretation

```{r cp-table, fig.cap="Cross-validated accuracy rate for the 20 different $\\alpha$ parameter values in our grid search. Lower $\\alpha$ values (deeper trees) help to minimize errors.", fig.height=3}

# caret cross validation results
mdl_caret <- train(
  present ~ .,
  data       = d_train,
  method     = "rpart",
  trControl  = trainControl(method = "cv", number = 10),
  tuneLength = 20)

ggplot(mdl_caret)
```

```{r dt-vip, fig.height=5.5, fig.cap="Variable importance based on the total reduction in MSE for the Ames Housing decision tree."}
vip(mdl_caret, num_features = 40, bar = FALSE)
```
**Question**: what are the top 3 most important variables of your model?
Latitude, WC_alt, and longitude

```{r dt-pdp, fig.width=10, fig.height= 3.5, fig.cap="Partial dependence plots to understand the relationship between lat, WC_alt and present."}
# Construct partial dependence plots
p1 <- partial(mdl_caret, pred.var = "lat") %>% autoplot()
p2 <- partial(mdl_caret, pred.var = "WC_alt") %>% autoplot()
p3 <- partial(mdl_caret, pred.var = c("lat", "WC_alt")) %>% 
  plotPartial(levelplot = FALSE, zlab = "yhat", drape = TRUE, 
              colorkey = TRUE, screen = list(z = -20, x = -60))

# Display plots side by side
gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

# Random Forests

## Fit

```{r out-of-box-rf}
# number of features
n_features <- length(setdiff(names(d_train), "present"))

# fit a default random forest model
mdl_rf <- ranger(present ~ ., data = d_train)

# get out of the box RMSE
(default_rmse <- sqrt(mdl_rf$prediction.error))
```

## Feature interpretation

```{r feature-importance}
# re-run model with impurity-based variable importance
mdl_impurity <- ranger(
  present ~ ., data = d_train,
  importance = "impurity")

# re-run model with permutation-based variable importance
mdl_permutation <- ranger(
  present ~ ., data = d_train,
  importance = "permutation")
```

```{r feature-importance-plot, fig.cap="Most important variables based on impurity (left) and permutation (right).", fig.height=4.5, fig.width=10}
p1 <- vip::vip(mdl_impurity, bar = FALSE)
p2 <- vip::vip(mdl_permutation, bar = FALSE)

gridExtra::grid.arrange(p1, p2, nrow = 1)
```
**Question**: How might variable importance differ between rpart and RandomForest in your model outputs?
